Concurrent Programming in Harmony

### 2021-12-03 book/paper.tex
Fixed bug

The first programming book I read cover to cover was \emph{The C Programming
Language} (first edition) by Brian W. Kernighan and Dennis M. Ritchie, which
was around 1980.  I did not know at the time that 10 years later Dennis,
the designer of the C programming language, would be my boss at AT\&T Bell Labs
in Murray Hill, NJ, while Brian would be my colleague in the same lab.
The first C program in the book printed
the string ``hello, world''.  Since then, most programming tutorials for
pretty much any programming language start with that example.

@@

\index{choose operator}%
But programs can usually have more than one execution and produce multiple
different outputs as a result.  This is usually as a result of different
inputs, but Harmony programs do not have inputs. Instead,
\autoref{fig:choice} demonstrates
\emph{nondetermistic choice} in Harmony programs.  In this case, the
program chooses to either print ``hello'' or ``world''.  The corresponding
DFA captures both possibilities.
You can think of the \texttt{choose} operator as enumerating all the possible
inputs to the program.

@@

\autoref{fig:threads} demonstrates \emph{methods} and \emph{threads} in Harmony.
In \autoref{fig:threads}, the code simply prints the string.
strings ``hello'' and ``world''.  Notice that this leads to an intermediate
state after ``hello'' is printed but before ``world'' is.  However, there
is still only one execution possible. \autoref{fig:threads}(b) shows two
threads, one printing ``hello'' and one printing ``world''.  Because the
threads run concurrently, the program can either output ``hello world'' or
``world hello''.  Printing in Harmony is atomic, so ``hweolrllod'' is not
a possible  output.

@@

\autoref{fig:triangle} shows another example of a Harmony program.
The example is a sequential program
and has a method \texttt{triangle} that takes
an integer number as argument.  Each method has a variable called
\textit{result} that eventually contains the result of the
method (there is no \textbf{return} statement in Harmony).  The method
also has a bound variable called \textit{n} containing the value of the
argument.  The \{ \textit{x..y} \} notation generates a set containing the numbers
from~\textit{x} to~\textit{y} (inclusive).
(Harmony does not have iterators and in particular does
not have a \texttt{range} operator.)
The last two lines in the program are
the most interesting.
The first assigns to \textit{x} some unspecified value in the range \texttt{0..N}
and the second verifies that $\mathtt{triangle}(x)$ equals $x(x+1)/2$.

@@

\begin{problems}
\item Harmony programs can usually be easily translated into
Python by hand.  For example,
\autoref{fig:incpy} is a Python version of \autoref{fig:inc}.
\begin{enumerate}
\item Run \autoref{fig:incpy} using Python.  Does the assertion fail?
\item Using a script, run \autoref{fig:incpy} 1000 times.
For example, if you are using the bash shell (in Linux or Mac OS X, say), you can do the following:
\begin{code}
\begin{verbatim}
for i in {1..1000}
do
    python Up.py
done
\end{verbatim}
\end{code}
If you're using Windows, the following batch script does the trick:
\begin{code}
\begin{verbatim}
FOR /L %%i IN (1, 1, 1000) DO python Up.py
PAUSE
\end{verbatim}
\end{code}
How many times does the assertion fail (if any)?
\end{enumerate}
\item \autoref{fig:incmany} is a version of \autoref{fig:incpy} that has each
incrementer thread increment \textit{count} \texttt{N} times.  Run \autoref{fig:incmany}
10 times (using Python).
Report how many times the assertion fails and what the value of \textit{count}
was for each of the failed runs.
Also experiment with lower values of \texttt{N}.
How large does \texttt{N} need to be for assertions to fail?
(Try powers of 10 for \texttt{N}.)
\item Can you think of a fix to \autoref{fig:inc}?  Try one or two different fixes
and run them through Harmony.  Do not worry about having to come up with a correct fix at this
time---the important thing is to develop an understanding of concurrency.
(Also, you do not get to use a \emph{lock}, yet.)
\end{problems}

>>>>

The first programming book that I read cover to cover was \emph{The C Programming
Language} (first edition) by Brian W. Kernighan and Dennis M. Ritchie, which
was around 1980.  I did not know at the time that 10 years later Dennis,
the designer of the C programming language, would be my boss at AT\&T Bell Labs
in Murray Hill, NJ, while Brian would be my colleague in the same lab.
The first C program in the book printed
the string ``hello, world''.  Since then, most programming tutorials for
pretty much any programming language start with that example.

@@

\index{choose operator}%
But programs can usually have more than one execution and produce multiple
different outputs as a result.  This is usually as a result of different
inputs, but Harmony programs do not have inputs. Instead,
\autoref{fig:choice} demonstrates
\emph{nondetermistic choice} in Harmony programs.  In this case, the
program chooses to either print ``hello'' or ``world''.  The corresponding
DFA captures both possibilities.
You can think of the \texttt{choose} operator as enumerating all possible
inputs to the program.

@@

\autoref{fig:threads} demonstrates \emph{methods} and \emph{threads} in Harmony.
In \autoref{fig:threads}(a), the code simply prints the
strings ``hello'' and ``world'', in that order.
Notice that this leads to an intermediate
state after ``hello'' is printed but before ``world'' is.  However, there
is still only one execution possible. \autoref{fig:threads}(b) shows two
threads, one printing ``hello'' and one printing ``world''.  Because the
threads run concurrently, the program can either output ``hello world'' or
``world hello''.  Printing in Harmony is atomic, so ``hweolrllod'' is not
a possible  output.

@@

\autoref{fig:triangle} shows another example of a Harmony program.
The example is a sequential program
and has a method \texttt{triangle} that takes
an integer number as argument.  Each method has a variable called
\textit{result} that eventually contains the result of the
method (there is no \textbf{return} statement in Harmony).  The method
also has a bound variable called \textit{n} containing the value of the
argument.  The \{ \textit{x..y} \} notation generates a set containing the numbers
from~\textit{x} to~\textit{y} (inclusive).
(Harmony does not have iterators and in particular does
not have a \texttt{range} operator like Python.)
The last two lines in the program are
the most interesting.
The first assigns to \textit{x} some unspecified value in the range \texttt{0..N}
and the second verifies that $\mathtt{triangle}(x)$ equals $x(x+1)/2$.

@@

\begin{problems}
\item Harmony programs can usually be easily translated into
Python by hand.  For example,
\autoref{fig:incpy} is a Python version of \autoref{fig:inc}.
\begin{enumerate}
\item Run \autoref{fig:incpy} using Python.  Does the assertion fail?
\item Using a script, run \autoref{fig:incpy} 1000 times.
For example, if you are using the bash shell (in Linux or Mac OS X, say), you can do the following:
\begin{code}
\begin{verbatim}
for i in {1..1000}
do
    python Up.py
done
\end{verbatim}
\end{code}
If you're using Windows, the following batch script does the trick:
\begin{code}
\begin{verbatim}
FOR /L %%i IN (1, 1, 1000) DO python Up.py
PAUSE
\end{verbatim}
\end{code}
How many times does the assertion fail (if any)?
\end{enumerate}
\item \autoref{fig:incmany} is a version of \autoref{fig:incpy} that has each
incrementer thread increment \textit{count} \texttt{N} times.  Run \autoref{fig:incmany}
10 times (using Python).
Report how many times the assertion fails and what the value of \textit{count}
was for each of the failed runs.
Also experiment with lower values of \texttt{N}.
How large does \texttt{N} need to be for assertions to fail?
(Try powers of 10 for \texttt{N}.)
\item Can you think of a fix to \autoref{fig:inc}?  Try one or two different fixes
and run them through Harmony.  Do not worry about having to come up with a correct fix at this
time---the important thing is to develop an understanding of concurrency.
(Also, you do not get to use the \texttt{atomically} keyword
or a \emph{lock}, yet.)
\end{problems}
